---
title: "Dealing with DSV files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dealing with DSV files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

"Delimiter-separated values" (DSV) files are (unfortunately) still a very common way
to serialize tabular data and are widely used as a data exchange format.
Since there isn't a strict technical standard that specifies how different data types
should be represented in plain text, and a lot of possible customizations are left to
the final users, it is very common to incurr in issues when data is imported from this
kind of files. 
The purpose of this article is to show some problems you can encounter 
loading data in a SQLite database from DSV files and how to deal with them.

## What to check in a DSV file

```{r setup}
library(RSQLite.toolkit)
```

Let's start with a simple example using real world data. To retrieve the examples
data, we can use the package `package?piggyback` to download some public datasets 
that has been stored in the [RSQLite.toolkit-tests](https://github.com/fab-algo/RSQLite.toolkit-tests)
GitHub repository[^1].

This example will deal with a table contained in a "standard" CSV file
(i.e. a text file with the first line containing column names, using the
comma to separate fields' values and the dot as a decimal separator for 
numbers); we will load  the data in a new table inside a SQLite database
using the `dbTableFromDSV()` function.

First we download the data from the repo:

```{r getdata}
library("piggyback", quietly = TRUE)
pb_download(file = "DOSE_V2.10.zip", dest = tempdir(),
            repo = "fab-algo/RSQLite.toolkit-tests", tag = "latest")

unzip(zipfile = file.path(tempdir(), "DOSE_V2.10.zip"), exdir = tempdir())
dir(file.path(tempdir(), "DOSE_V2.10"))

data_file <- file.path(tempdir(), "DOSE_V2.10/DOSE_V2.10.csv")
```

After unzipping the downloaded file, we get the `DOSE_V2.10.csv` file that
contains the dataset we are interested in.
We can now check the number of rows in the input file:

```{r inspect_file2}
n_rows <- length(count.fields(data_file))
n_rows
```

So we expect to have `r format(n_rows-1, nsmall=0)` records in the SQLite table.
Now we connect to a sample database (i.e. `tests.sqlite`): 

```{r dbcon}
dbcon <- dbConnect(dbDriver("SQLite"), file.path(tempdir(), "tests.sqlite"))
```
and move the dataset in a table there, calling the `dbTableFromDSV()` function 
specifying the connection to the database (i.e.  `dbcon` parameter) and the 
table name (i.e. `table_name` parameter). We don't need to specify the parameters
for the CSV file strucuture (i.e. `header = TRUE`, `sep = ","`, `dec = "."`, 
`grp = ""`) since the default values will be fine.

```{r, label=loaddata1, error=TRUE}
## do not run: error
dbTableFromDSV(input_file = data_file, dbcon = dbcon, table_name = "DOSE")
```

To understand what happened we need to keep in mind that the `dbTableFromDSV()` 
function calls `file_schema_dsv` function, that uses the *`read.table`* function 
(from the `utils` package) to guess the table structure stored in the DSV file, 
and than the *`scan`* function to actually read the data in a data frame and 
write it to the SQLite table.

Both these functions make some assumptions on how the data is stored in the DSV
file as "plain text". More in detail, both functions use the parameters listed in the 
following table along with their default values.

| parameters   | scan    | read.table |
|--------------|---------|------------|
| quote        | `"'\""` | `"\"'"`    |
| allowEscapes | `FALSE` | `FALSE`    |
| skipNul      | `FALSE` | `FALSE`    |
| strip.white  | `FALSE` | `FALSE`    |
| comment.char | `""`    | `"#"`      |
| na.strings   | `"NA"`  | `"NA"`     |
| fill         | `FALSE` | `FALSE`    |
| fileEncoding | `""`    | `""`       |

:`scan` and `read.table` parameters to handle data interpretation.

These parameters control the conventions used to interpret the text read 
from the file:

- *quoting*: convention used to inclued in strings "special" characters or to
   preserve the exact string without interpretation.
   Parameter: `quote`
- *escaping*: rule used to represent characters that would otherwise be interpreted 
   as delimiters or special commands.
   Parameter: `allowEscapes`
- *NUL characters*: how to manage ASCII nul(s) characters that could be found 
   inside strings.
   Parameter: `skipNul`
- *whitespace stripping*: if unwanted spaces from the beginning, end, or middle of a 
   string should be removed. 
   Parameter: `strip.white`
- *commenting*: character used to identify that one line (or the rest of a line after
   the "comment" identifier) does not contain data and should be ignored.
   Parameter: `comment.char`
- *missing data*: how missing data is represented in the file.
   Parameter: `na.strings`
- *missing fields*: how to manage rows with less fields then expected.
   Parameter: `fill`
- *encoding*: the technical standard used to represent characters in binary form, as
   for example ASCII, UTF-8, EBCDIC.
   Parameter: `fileEncoding`

To understand how to use each of these parameters you are strongly encouraged to read 
the help pages of the `scan` and `read.table` functions.

It should be noted that the default values are seldom (... almost never) the right
ones and we should invest some time to better understand the strategies used by the
data source. If we have a technical description of the assumptions used in 
creating the file we can follow those guidelines, otherwise we have to inspect the 
file and determine them by direct guess.

To explore a DSV file (especially big files) it is better to use log file explorers
and avoid general purpose text editors. As an example, we can use [klogg](https://klogg.filimonov.dev/) 
that is available for all major operating systems.

If we open the file and inspect some records we understand that:

- there are strings with comma and single quote inside, surrounded by 
  double quotes, so the file is using a specific quoting convention;
- missing data is represented by a "zero-lenght" string;
- many strings use special characters that need a specific enconding 
  system; in this file the UTF-8 enconding is used.

For the input file used in this example we need:

- to explicitly force the `quote` parameter to `"\""` (i.e. only the double quote 
  character) and the `na.strings` parameter to `""`;
- to turn off the commenting option: `comment.char = ""`;
- to tell to the `base::scan()` function that the input file is encoded in UTF-8: 
  `fileEncoding = "UTF-8"`.

```{r schema}
schema <- file_schema_dsv(input_file = data_file,
                          quote = "\"", na.strings = "",
                          comment.char = "", fileEncoding = "UTF-8")
```

```{r loaddata2}
dbTableFromDSV(input_file = data_file, dbcon = dbcon, table_name = "DOSE",
               drop_table = TRUE, quote = "\"", na.strings = "",
               comment.char = "", fileEncoding = "UTF-8")
```

As we can see now we got all the expected records in the SQLite table.

```{r close_db}
dbDisconnect(dbcon)
```

## Issues with column names

Unfortunately, when dealing with DSV files, problems are not limited to data interpretation
only. Another common source of issues are the column names used in the input file.

When working with DSV files, column names can present several challenges.
While DSV files may contain column headers that are valid in their original context
(such as Excel or other applications), these names might not conform to SQLite's naming
conventions or best practices.

SQLite has specific rules and restrictions regarding column names, including:

- Must begin with a letter (A-Z, a-z) or an underscore (_)
- Can contain letters, digits (0-9), and underscores (_)
- Cannot be a reserved keyword in SQLite
- Should avoid special characters and spaces for better compatibility

You can find more details about SQLite's naming conventions in the
[official documentation](https://www.sqlite.org/lang_keywords.html) and 
[CREATE TABLE](https://www.sqlite.org/lang_createtable.html) page. There are
also some interesting discussions on Stack Overflow:

- [What are valid table names in SQLite?](https://stackoverflow.com/questions/3694276/what-are-valid-table-names-in-sqlite)
- [What SQLite column name can be/cannot be?](https://stackoverflow.com/questions/3373234/what-sqlite-column-name-can-be-cannot-be)
- [SQLite table and column name requirements](https://stackoverflow.com/questions/23770480/sqlite-table-and-column-name-requirements)


### Common issues with column names

Some typical problems you might encounter include:

- *Special characters*: Column names containing spaces, hyphens, or other 
  special characters (e.g., `"Patient-ID"`, `"Test Result"`, `"Cost($)"`)
- *Reserved keywords*: Column names that match SQLite reserved words
  (e.g., `"SELECT"`, `"TABLE"`, `"WHERE"`)
- *Starting with numbers*: Column names beginning with digits 
  (e.g., `"2023_Sales"`)
- *Case sensitivity*: While SQLite is case-insensitive for column names,
  mixing cases can lead to confusion
- *Empty or duplicate names*: Missing headers or columns with identical names

You should always inspect the column names in your DSV files before importing them
into a SQLite database to ensure they meet these criteria. Failing to do so can lead
to errors during the import process or unexpected behavior when querying the database.


### How RSQLite.toolkit handles column names

The `dbTableFromDSV()` function automatically sanitizes column names to ensure compatibility with SQLite by:

1. Replacing spaces and special characters with underscores
2. Adding a prefix to column names that start with numbers
3. Handling reserved keywords by modifying them
4. Ensuring uniqueness by appending suffixes to duplicate names

### Example

Let's create a simple example to demonstrate this behavior:


## Conclusions

To handle in a safer way DSV files (and avoid some unexpected results)

	  



[^1]: The original source of each dataset is described in the README page of the GitHub repo.

